\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-08 paper/ms.tex
words about semiseparable matrices

In this paper, we present a method for directly and exactly computing a class
of GP models that scales linearly with the number of data points
$\mathcal{O}(N)$ for one dimensional data sets.
\response{Unlike earlier linear methods using Kalman filters \citep[for
example,][]{Kelly:2014}, this novel algorithm exploits the semi-separable
structure of a specific class of covariance matrices to directly factorize and
solve the system.
}
This method can only be used with one-dimensional datasets and the covariance
function must be represented by a mixture of exponentials; we will return to a
discussion of what this means in detail in the following sections.
However, the measurements don't need to be evenly spaced and the uncertainties
can be heteroscedastic.

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
\response{The semi-separable nature of these matrices was first recognized by
\citet{Ambikasaran:2015}, building on intuition from a twenty year old paper
\citep{Rybicki:1995}.}
As we will discuss in the following pages, this choice of kernel function
arises naturally in some physical systems and we demonstrate it can be used as
an effective model in some other cases.
This method is especially appealing compared to other similar methods~--~we
discuss these comparisons in \sect{compare}~--~because it is exact, simple,
and fast.

@@

\section{The celerite model}

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
operations when the covariance function is given by
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})
\end{eqnarray}
\response{by recognizing that the matrix $K_\bvec{\alpha}$ generated by this
kernel is semi-separable.}
In this case, the inverse is dense but \eq{gp-likelihood} can still be
evaluated with a scaling of $\mathcal{O}(N\,J^2)$, where $J$ is the number of
components in the mixture.
\response{In the following section, we derive a novel method with the same
$\mathcal{O}(N\,J^2)$ scaling that can be used when
$k_\bvec{\alpha}(\tau_{nm})$ is positive semi-definite.}

@@

We now turn to a discussion of \celeriteterm\ as a model of astrophysical
variability.
A common concern in the context of GP modeling in astrophysics is the lack of
physical motivation for the choice of kernel functions.
Kernels are often chosen simply because they are popular with little
consideration of the impact of this decision.
While this isn't necessarily a problem for the inferred results, in this
section we discuss an exact physical interpretation of the \celeriteterm\
kernel that will be applicable to many astrophysical systems but especially
the time-variability of stars.

@@

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations and we return to give some
specific examples in \sect{examples-real}.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
of stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \celeriteterm\ method
described in the previous section.

@@

\section{Semi-separable matrix factorization}

The relationship between the \celeriteterm\ model and semi-separable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015}.
A rank-$R$ semi-separable matrix is a matrix $K$ where the elements can be
written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.  Note that the last two terms in this equation are
transposes of one another.
For the \celeriteterm\ model described in the previous section, the rank of
the semi-separable matrix is $R = 2\,J$ and the components of the relevant
matrices are
\begin{eqnarray}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}

Note that rather than storing the full $N \times N$ covariance matrix, it
is only necessary to store the $2J \times N$ matrices $U$ and $V$, as well
as the diagonal terms, for a total storage volume of $(4J+1) \times N$ floating
point numbers.

@@

To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
compute the log-determinant of $K$ and solve linear systems of the form
$K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is a rank-$2\,J$
semi-separable matrix.
If $K$ is positive semi-definite, these can be computed using the Cholesky
factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,W^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
$N \times 2\,J$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,W^\T)\,D
    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $W$.
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,W_{n-1,j}\,W_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
observation~--~so the total cost of this factorization scales as
$\mathcal{O}(N\,J^2)$.

Note that if some \celeriteterm\ terms are real, with $b_j=d_j=0$,
then these terms only add one to the rank of $U$ and $V$.  We use this fact
to simplify the algebra when real components are included:  if $J_{real}$
and $J_{complex}$ terms are included, with $J_{real}+J_{complex}=J$, then
the $U$, $V$, and $W$ matrices have size $N \times (J_{real}+2J_{complex})$.

\subsection{Pre-conditioning}

Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
overflow and underflow issues in many realistic cases because both $U$ and $V$
have factors of the form $e^{\pm c_j t}$, where $t$ can be any arbitrarily large(small)
real number;  these factors can cancel in the end, but will not if over(under)flow
is an issue.  This can be avoided with the appropriate reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \\
\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
\end{eqnarray}
and the set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}

@@

\section{An alternative solver using the extended matrix formalism}

@@

\section{Semi-separable operations}

To help clarify these equations, we write out here an example of a semi-separable matrix with a

@@

Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(JN)$
operations with a stable recursive algorithm using the semi-separable form:
\begin{equation}
\bvec{y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) \bvec{z}.
\end{equation}
The algorithm for multiplication is given by
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \sum_{j=1}^J \tilde U_{n+1,j} z_{n+1}\right)\\
f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \sum_{j=1}^J \tilde V_{n-1,j} z_{n-1}\right)\\
y_n &=& A_{n,n} z_n + \sum_{j=1}^J \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,J$.  The recursion requires
two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards
from $n=2$ to $n=N$ to compute $f^-_{n,j}$.  \dfmtodo{I'm getting confused by zero-based
indexing in your C code...  this may be worth scrutinizing to see if I screwed up.}

@@

\section{Ensuring positive definiteness}

>>>>

In this paper, we present a method for directly and exactly computing a class
of GP models that scales linearly with the number of data points
$\mathcal{O}(N)$ for one dimensional data sets.
\response{Unlike earlier linear methods using Kalman filters \citep[for
example,][]{Kelly:2014}, this novel algorithm exploits the semiseparable
structure of a specific class of covariance matrices to directly factorize and
solve the system.
}
This method can only be used with one-dimensional datasets and the covariance
function must be represented by a mixture of exponentials; we will return to a
discussion of what this means in detail in the following sections.
However, the measurements don't need to be evenly spaced and the uncertainties
can be heteroscedastic.

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
\response{The semiseparable nature of these matrices was first recognized by
\citet{Ambikasaran:2015}, building on intuition from a twenty year old paper
\citep{Rybicki:1995}.}
As we will discuss in the following pages, this choice of kernel function
arises naturally in some physical systems and we demonstrate it can be used as
an effective model in some other cases.
This method is especially appealing compared to other similar methods~--~we
discuss these comparisons in \sect{compare}~--~because it is exact, simple,
and fast.

@@

\section{The celerite model}\sectlabel{celerite}

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
operations when the covariance function is given by
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})
\end{eqnarray}
\response{by recognizing that the matrix $K_\bvec{\alpha}$ generated by this
kernel is semiseparable.}
In this case, the inverse is dense but \eq{gp-likelihood} can still be
evaluated with a scaling of $\mathcal{O}(N\,J^2)$, where $J$ is the number of
components in the mixture.
\response{In \sect{factor}, we derive a novel method with the same
$\mathcal{O}(N\,J^2)$ scaling that can be used when
$k_\bvec{\alpha}(\tau_{nm})$ is positive semi-definite.}

@@

We now turn to a discussion of \celeriteterm\ as a model of astrophysical
variability.
A common concern in the context of GP modeling in astrophysics is the lack of
physical motivation for the choice of kernel functions.
Kernels are often chosen simply because they are popular, with little
consideration of the impact of this decision.
While this doesn't necessarily impact the inferred results, in this section we
discuss an exact physical interpretation of the \celeriteterm\ kernel that
will be applicable to many astrophysical systems but especially the
time-variability of stars.

@@

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations and we return to give some
specific examples in \sect{examples}.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
of stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \celeriteterm\ method
described in the next section.

@@

\section{Semiseparable matrices \& celerite}\sectlabel{factor}

The previous two sections describe the \celeriteterm\ model and its physical
interpretation.
In this section, we derive a new scalable direct solver for covariance
matrices generated by this model to compute the GP likelihood
(\eqalt{gp-likelihood}) for large datasets.
This method exploits the semiseparable structure of these matrices to compute
their Cholesky factorization in $\mathcal{O}(N\,J^2)$ operations.

The relationship between the \celeriteterm\ model and semiseparable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015} building
on earlier work by Rybicki \& Press \citep{Rybicki:1995}.
\citet{Ambikasaran:2015} derived a scalable direct solver for any general
semiseparable matrix and applied this method to \celeriteterm\ models with
$b_j = 0$ and $d_j = 0$.
In \app{extended}, we generalize this earlier method to include non-zero $b_j$
and $d_j$.
It turns out, however, that we can derive a higher performance algorithm by
only considering positive semi-definite (PSD) matrices.
In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
PSD.

To start, we define a rank-$R$ semiseparable matrix as a matrix $K$ where the
elements can be written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.
In this representation, rather than storing the full $N \times N$ matrix $K$,
it is sufficient to only store the $N \times R$ matrices $U$ and $V$, and the
$N$ diagonal entries of $A$, for a total storage volume of $(2\,R+1)\,N$
floating point numbers.
Given a semiseparable matrix of this form, it is known that matrix-vector
products and matrix-matrix products can be evaluated in $\mathcal{O}(N)$
operations and the interested reader is directed to \citet{Vandebril:2007} for
more details about semiseparable matrices in general.

@@

For our purposes, we want to compute the log-determinant of $K$ and solve
linear systems of the form $K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is
a rank-$R$ semiseparable PSD matrix.
These can both be computed using the Cholesky factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,W^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
$N \times R$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,W^\T)\,D
    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $W$.
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,W_{n-1,j}\,W_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{R}\sum_{k=1}^{R} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{R}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $R \times R$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(R^2)$ and this must be run $N$ times~--~once for each row~--~so
the total cost of this factorization scales as $\mathcal{O}(N\,R^2)$.

\subsection{Scalable factorization of the celerite model}

The \celeriteterm\ model described in \sect{celerite} can be exactly
represented as a rank $R = 2\,J$ semiseparable matrix\footnote{We note that if
some \celeriteterm\ terms are real ($b_j=0$ and $d_j=0$) then these terms only
add one to the rank.
If a \celeriteterm\ model has $J_\mathrm{real}$ real and $J_\mathrm{complex}$
general terms, the semiseparable rank is only $J_\mathrm{real} +
2\,J_\mathrm{complex}$.}
where the components of the relevant matrices are
\begin{eqnarray}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}
Na\"ive implementation of the algorithm in \eq{algorithm} to this system will
result in numerical overflow and underflow issues in many realistic cases
because both $U$ and $V$ have factors of the form $e^{\pm c_j t}$, where $t$
can be any arbitrarily large real number.
This issue was also noted by \citet{Ambikasaran:2015} and it can be avoided
with a reparameterization similar to the one they use.
We define the following pre-conditioned variables
\begin{eqnarray}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \\
\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
\end{eqnarray}
and the set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm to compute the Cholesky factorization of a \celeriteterm\ model
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}
As before, a single iteration of this algorithm requires $\mathcal{O}(J^2)$
operations and it is repeated $N$ times for a total scaling of
$\mathcal{O}(N\,J^2)$.
Furthermore, $\tilde{V}_{n,j}$ and $A_{n,n}$ can be updated in place and used
to represent $\tilde{W}_{n,j}$ and $D_{n,n}$, giving a total memory footprint
of $(6\,J + 1)\,N + J\,(J-1)/2$ floating point numbers.

@@

\section{An alternative solver using the extended matrix
formalism}\sectlabel{extended}

@@

\section{Semiseparable operations}

To help clarify these equations, we write out here an example of a semiseparable matrix with a

@@

Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(JN)$
operations with a stable recursive algorithm using the semiseparable form:
\begin{equation}
\bvec{y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) \bvec{z}.
\end{equation}
The algorithm for multiplication is given by
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \sum_{j=1}^J \tilde U_{n+1,j} z_{n+1}\right)\\
f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \sum_{j=1}^J \tilde V_{n-1,j} z_{n-1}\right)\\
y_n &=& A_{n,n} z_n + \sum_{j=1}^J \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,J$.  The recursion requires
two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards
from $n=2$ to $n=N$ to compute $f^-_{n,j}$.  \dfmtodo{I'm getting confused by zero-based
indexing in your C code...  this may be worth scrutinizing to see if I screwed up.}

@@

\section{Ensuring positive definiteness}\sectlabel{psd}
